Activity Monitoring Report
-----------------------------------------


####Phase I--Loading and preprocessing the data.

Step 1: Load the data into R

```{r echo=TRUE}
activity<-read.csv('./data/activity.csv', header=TRUE, na.strings='NA', stringsAsFactor=FALSE)
```
Step 2: Determine the number of dimensions in the dataset.
```{r ech=TRUE}
dim(activity)
```
Step 3: Examine the structure of the dataset.
```{r echo=TRUE}
str(activity)
```
Step 4: Examine the head and tail of the dataset.
```{r echo=TRUE}
head(activity, 20)
tail(activity, 20)
```
Step 5: Transform the 'date' variable to date mode
```{r echo=TRUE}
activity<-transform(activity,
                    date = as.Date(date, format='%Y-%m-%d')
                    )
```
Step 6: Summarize each variable in the dataset.
```{r echo=TRUE}
summary(activity)
```
Step 7: Review the names of each variable.
```{r echo=TRUE}
names(activity)
```
Analysis Phase:
What is the mean number of steps taken per day?
```{r echo=TRUE}
library(sqldf)
avg_num_steps_per_day<-sqldf('SELECT date, avg(steps) as avg_steps FROM activity 
                             GROUP BY 
                             Date 
                             ORDER BY 
                             date;')
avg_num_steps_per_day

with(activity, aggregate(steps, by=list(date=date), FUN=mean))
```
If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
```{r echo=TRUE, fig.width=12}
total_steps_per_day<-sqldf('SELECT date, sum(steps) as daily_total_steps FROM activity
                          GROUP BY
                          date
                          ORDER BY
                          date;')
str(total_steps_per_day)

total_steps_per_day<-transform(total_steps_per_day,
                               daily_total_steps = as.numeric(daily_total_steps)
                               )

min.step_day<-min(total_steps_per_day$daily_total_steps, na.rm=TRUE)

max.step_day<-max(total_steps_per_day$daily_total_steps, na.rm=TRUE)



with(total_steps_per_day, plot(daily_total_steps ~ date,
                               type = 'h',
                               lwd = 4,
                               axes=FALSE,
                               ylab = 'Total Steps',
                               xlab = 'Calendar Date',
                               main = 'Total Number Of Steps Taken Each Day\nHistogram'
                               )
     )
box()
axis(1, total_steps_per_day$date, format(total_steps_per_day$date, format='%m-%d'), las=2 )
axis(2, ylim=c(0, max.step_day), las =1)
```
Calculate and report the mean and median of the total number of steps taken per day
```{r echo=TRUE}
mean_steps_day<-round(mean(total_steps_per_day$daily_total_steps, na.rm=TRUE), 2)
mean_steps_day<-as.character(mean_steps_day)
```
The mean number of steps taken per day is `r mean_steps_day` 

```{r echo =TRUE}
median_steps_day<-median(total_steps_per_day$daily_total_steps, na.rm=TRUE)
steps_day_median<-as.character(median_steps_day)
```
The median number of steps taken per day is `r steps_day_median`

What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, lineplot, fig.height= 8, fig.width = 12, echo =TRUE}

interval_avg_steps<-sqldf('SELECT interval, avg(steps) as avg_steps_by_interval FROM activity
                        GROUP by
                        interval
                        ORDER BY 
                        interval;')

with(interval_avg_steps, plot(avg_steps_by_interval ~ interval,
                              type = 'l',
                              col = 'red',
                              ylab ='Avg Number of Steps',
                              main = 'Average Daily Activity Pattern'
                              )
     )


```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
avg_num_steps_per_day<-na.omit(avg_num_steps_per_day)

avg_num_steps_per_day<-transform(avg_num_steps_per_day,
                                 avg_steps = as.numeric(avg_steps)
                                 )

max_steps_per_day<-max(avg_num_steps_per_day$avg_steps)

max_steps_per_day

which_date<-avg_num_steps_per_day[which(avg_num_steps_per_day$avg_steps==max_steps_per_day), 1]

```
The  5-minute interval, on average across all the days in the dataset, contains the maximum number of steps is `r which_date`


Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r echo=TRUE}
na.activity<-activity[!complete.cases(activity),]
total_number_missing_values<-nrow(na.activity)

```
The total number of missing values in the dataset is `r total_number_missing_values`

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r echo = TRUE}

activity_fill<-sqldf("SELECT A.date, A.interval, A.steps, I.avg_steps_by_interval FROM activity as A
                       LEFT OUTER JOIN 
                               interval_avg_steps AS I
                                        ON
                                         A.interval = I.interval ;")

str(activity_fill)

activity_fill<-transform(activity_fill,
                         steps = as.numeric(steps)
                         )

str(activity_fill)

num_records<-nrow(activity_fill)
num_records
for(i in 1:num_records){
        if( is.na(activity_fill$steps[i])==TRUE){
              activity_fill$steps[i]<-activity_fill$avg_steps_by_interval[i]
      }

}

activity_fill<-activity_fill[,1:3]

head(activity_fill, 10)

summary(activity_fill$steps)

activity_agg<-sqldf('SELECT date, sum(steps) as total_steps FROM activity_fill
                        GROUP BY date
                        ORDER BY date;')
```

Make a histogram of the total number of steps taken each day.


```{r echo=TRUE, fig.width= 12}
with(activity_agg, plot(total_steps ~ date,
                                type = 'h',
                                axes = FALSE,
                                ylab = 'Total Steps')
     )
box()
axis(1, activity_agg$date, format(activity_agg$date, '%m-%d'), las=2)
axis(2, ylim=c(0, max(activity_agg$steps)))
```

Calculate and report the mean and median total number of steps taken per day

```{r echo=TRUE}
activity_fill_mean<-mean(activity_agg$total_steps)

activity_fill_mean<-as.character(round((activity_fill_mean),2))

activity_fill_median<-median(activity_agg$total_steps)

activity_fill_median<-as.character(round((activity_fill_median),2))

```
The new mean is `r activity_fill_mean` 

The new median is `r activity_fill_median`

 Do these values differ from the estimates from the first part of the assignment? 
 
 Yes. the values differ from the original dataset. The origial dataset mean and means were `r mean_steps_day` and `r steps_day_median` 
 
```{r echo=TRUE}
activity_fill_mean<-as.numeric(activity_fill_mean)

mean_steps_day<-as.numeric(mean_steps_day)

activity_mean_delta<-(activity_fill_mean - mean_steps_day)/mean_steps_day

activity_mean_delta

activity_fill_median<-as.numeric(activity_fill_median)

steps_day_median<-as.numeric(steps_day_median)

activity_media_delta<-(activity_fill_median - steps_day_median)/steps_day_median

activity_media_delta

```
 What is the impact of imputing missing data on the estimates of the total daily number of steps?
 
 Insignificant.  The difference between the original and the new means and medians are 0.0%. As shown in the summaries below, the replacement of missing values with average values for each interval did not impact the results
 
```{r echo=TRUE}
summary(activity_fill$steps)

summary(activity$steps)

```

```{r echo=TRUE}
activity_fill$weekpart<-NA

num_rec<-nrow(activity_fill)

for(i in 1:num_rec){
        if(weekdays(activity_fill$date[i])=='Saturday'){
                activity_fill$weekpart[i]<-'weekend'
        } else if (weekdays(activity_fill$date[i])=='Sunday'){
                activity_fill$weekpart[i]<-'weekend'
        } else {
              activity_fill$weekpart[i]<-'weekday'  
        }
}


head(activity_fill, 100)

activity_fill<-transform(activity_fill,
                         weekpart=factor(weekpart)
                         )


summary(activity_fill$weekpart)

```
 
 Make a panel plot containing a time series plot 
 
```{r echo=TRUE}

weekdays_subset<-sqldf("SELECT interval, avg(steps) as avg_steps FROM activity_fill
                       WHERE
                        weekpart='weekday'
                       GROUP BY
                        interval;")

with(weekdays_subset, plot(avg_steps ~ interval,
                           type = 'l',
                           col = 'red',
                           ylab = 'Avg Number Steps',
                           main = 'Weekday Activity Monitor'
                           )
     )
     
weekend_subset<-sqldf("SELECT interval, avg(steps) as avg_steps FROM activity_fill
                       WHERE
                        weekpart='weekend'
                       GROUP BY
                        interval;")


with(weekend_subset, plot(avg_steps ~ interval,
                           type = 'l',
                           col = 'blue',
                           ylab = 'Avg Number Steps',
                           main = 'Weekend Activity Monitor'
                           )
     )

```
 
 
 
 